---
title: R/lineup2 User Guide
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R/lineup2 User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
---

```{r knitr_options, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=4.5)
```


[R/lineup2](https://github.com/kbroman/lineup2)
is an [R](https://www.r-project.org) package
providing tools for detecting and
correcting sample mix-ups between two sets of measurements, such as
between gene expression data on two tissues. It's a revised
version of [lineup](https://github.com/kbroman/lineup), to be more
general and not so closely tied to the [R/qtl](http://rqtl.org)
package.

We will illustrate some of the tools in this User Guide. We will use
the example data included in the package `lineup2ex`. These are a
subset of data from [Broman et al.
(2015)](https://doi.org/10.1534/g3.115.019778) (see also [Tian et al.
2015](https://doi.org/10.1534/genetics.115.179432)), and concern gene
expression microarray data on about 500 mice in two tissues,
gastrocnemius muscle (abbreviated "gastroc") and pancreatic islets
(abbreviated "islet"). While the original data included 40,000 gene
expression measurements, the example data consider just a subset of
200 genes: 100 chosen to be highly correlated between the two tissues,
and 100 others chosen at random.

When we load the [lineup2](https://github.com/kbroman/lineup2)
package, the data are immediately available, as `lineup2ex`, which is
a list containing the two tissues, each of which is a matrix, samples
&times; genes.

```{r load_package}
library(lineup2)
gastroc <- lineup2ex$gastroc
islet <- lineup2ex$islet
```

We seek to identify sample mix-ups: mislabeled rows in one or the
other data sets. One approach to do this is to first identify a
subset of correlated columns, and then to look at the correlations
between rows, for that subset of columns.

### Aligning rows or columns

To calculate the correlations between columns, we first need to align
the rows between the two matrices. We can do this with the function
`align_matrix_rows()`. There is a similar function
`align_matrix_cols()` for aligning columns.

```{r align_rows}
aligned <- align_matrix_rows(gastroc, islet)
```

The original matrices have `r nrow(gastroc)` and `r nrow(islet)` rows.
The output of `align_matrix_rows()` is a list with two matrices, where
the original matrices are subsetted to their common rows, based on
their rownames. Here, each matrix in the output has `r nrow(aligned[[1]])`
rows, which is the number of samples in common between the two
tissues.

We could use `align_matrix_cols()` to align the columns (subsetting to
the common column names and putting them in the same order), but here they
are already aligned, and so no change occurs.

```{r align_cols}
aligned <- align_matrix_cols(aligned[[1]], aligned[[2]])
```

### Correlations between matrices

With the rows now aligned, we can now calculate the correlation in
gene expression between the two tissues, to identify genes that will
be useful for measuring sample similarity.

The function `corr_betw_matrices()` is used to calculate the
correlation between the columns in one matrix and the columns in
another matrix. There are several options; here we will use `what="paired"`
to get just the correlations between the paired columns.

```{r calc_paired_corr}
paired_corr <- corr_betw_matrices(aligned[[1]], aligned[[2]])
```

The result is a vector of `r ncol(aligned[[1]])` correlations, for the
pairs of columns. Here is a plot of the sorted correlations.

```{r to_reset_par, include=FALSE}
old_mar <- par("mar")
old_mfrow <- par("mfrow")
```

```{r plot_paired_corr}
par(mar=c(4.1, 5.1, 1.1, 1.1))
plot(sort(paired_corr), ylab="Correlations between column pairs, sorted", las=1, pch=16)
abline(h=0, lty=2, col="gray60")
```

Because of the way the genes were selected, there are 100 genes
with between-tissue correlation scattered around 0, and then another
100 genes with correlations > 0.75. We will grab that top 100 for
assessing sample similarity.

```{r select_top_genes}
selected_genes <- which(paired_corr > 0.75)
```

### Correlations/distances between samples

We can now calculate the similarity between the samples using these
selected genes. The simplest way to do so is by again using
`corr_betw_matrices()`, but with the transposed matrices so that the
samples become the columns.

```{r sample_correlations}
corr_betw_samples <- corr_betw_matrices(t(gastroc[,selected_genes]),
                                        t(islet[,selected_genes]), what="all")
```

The output is a `r nrow(gastroc)` &times; `r nrow(islet)` matrix, with
the (_i_,_j_)th element being the correlation between the _i_th row of
`gastroc` and the _j_th row of `islet`.

Note that the relate function `dist_betw_matrices()` can be used to
calculate the Euclidean or mean absolute distance between samples.
There is also `dist_betw_arrays()` for arrays with >2 dimensions.
Unlike `corr_betw_matrices()`, these functions look for distance
between rows not columns.


### Summarize correlations/distances

```{r hist_self_nonself, fig.height=6}
hist_self_nonself(corr_betw_samples)
```

```{r self_values}
self <- get_self(corr_betw_samples)
```

There is also a function `get_nonself()` for pulling out the non-self values.




### Identify sample mix-ups

```{r best_by_row_and_col}
best_byrow <- get_best(corr_betw_samples, get_min=FALSE, dimension="row")
best_bycol <- get_best(corr_betw_samples, get_min=FALSE, dimension="column")
```

```{r plot_self_v_best}
par(mfrow=c(2,1))
plot(self, best_byrow, xlab="Self-self distance", ylab="Distance to best islet")
plot(self, best_bycol, xlab="Self-self distance", ylab="Distance to best gastroc")
```

gastroc
(3655, 3659)

islet
(3598, 3599)
(3296 != 3295)





### Second-best samples

```{r second_best_by_row_and_col}
secbest_byrow <- get_2ndbest(corr_betw_samples, get_min=FALSE, dimension="row")
secbest_bycol <- get_2ndbest(corr_betw_samples, get_min=FALSE, dimension="column")
```




### References

Broman KW, Keller MP, Broman AT, Kendziorski C, Yandell BS, Sen Ś,
Attie AD (2015) Identification and correction of sample mix-ups in
expression genetic data: A case study. G3 5:2177-2186
[doi:10.1534/g3.115.019778](https://doi.org/10.1534/g3.115.019778)

Tian J, Keller MP, Oler AT, Rabaglia ME, Schueler KL, Stapleton DS,
Broman AT, Zhao W, Kendziorski C, Yandell BS, Hagenbuch B, Broman KW,
Attie AD (2015) Identification of the bile acid transporter Slco1a6 as
a candidate gene that broadly affects gene expression in mouse
pancreatic islets. Genetics 201:1253-1262
[doi:10.1534/genetics.115.179432](https://doi.org/10.1534/genetics.115.179432)


```{r reset_par, include=FALSE}
par(mar=old_mar, mfrow=old_mfrow)
```
